Using Scagnostics to find Interesting Projections of Multivariate Data

### 2018-09-06 scagnostics.Rmd
Adding measure descriptions to the introduction.

A large number of projection pursuit indices have been developed [@f87] [@CBC92] [@Lee05] [@AHC02] [@CEM:CEM2568] [@JS87] [@5508437]. Many of these are developed to detect departure from multivariate normal, which includes clusters or outliers, or nonlinearity. Some have been developed to detect separations between known groups. However, if one wants to find unusual shapes, or shape differences between groups there are no options. One reason, is that it is hard, because defining what is meant by shape is not straightforward. An approach that seems plausible begins with scagnostics, developed by @scag, and explored in @WW08.

*** Motivating example ??
This could be the pdf fit data, more structure and shape than other datasets we have?

>>>>

A large number of projection pursuit indices have been developed [@f87] [@CBC92] [@Lee05] [@AHC02] [@CEM:CEM2568] [@JS87] [@5508437]. Many of these are developed to detect departure from multivariate normal, which includes clusters or outliers, or nonlinearity. Some have been developed to detect separations between known groups. However, if one wants to find unusual shapes, or shape differences between groups there are no options. One reason, is that it is hard, because defining what is meant by shape is not straightforward.

A related question is the selection of pair-wise scatter plots in datasets with large numbers of parameters, for which different methods have been developed. One approch are scatterplot diagnostics, "scagnostics", developed by @scag, and explored in @WW08. They are calculated from a set of elementary graph-theoretic building blocks: the convex hull, the alpha hull and the minimal spanning tree (MST). These are used to define eight measures: Outlying, Skewed, Sparse, Clumpy, Striated, Convex, Skinny and Stringy. While in principle each measure could be used on its own to define a ranking, the original idea suggested considering pair-wise scatter plots of the scagnostics measures for identification of "characteristic" and "unusual" 2-d views in the original parameter space. Alternative methods suggested by (Katrins Thesis or maybe her R package?) suggest using parallel coordinates, glyphs or a so-called scagramm (based on the idea of corrgrams) that shouold be coupled to interactive brushing for selection of pair-wise scatter plots to be displayed.
(FIXME just give reference rather than all definitions here?)
(Issues with scagnostics already for selection of pairs have been described in Katrins Thesis, maybe have short summary of that here?)
Given the various shortcomings of the scagnostics approach we consider several other measures that can directly be interpreted as ranking of information contained in a given 2-d view.
Two such additional measures have been defined in @mbgraphic. The first is based on smoothing splines, and is defined such that it will take maximum values when the data distribution can be described by a functional form. For a scatterplot with variable X and Y we calculate it as
\begin{equation}
splines_{2d} = max(1-\frac{Var(res_{X\sim Y})}{Var(X)}, 1-\frac{Var(res_{Y\sim X})}{Var(Y)})
\end{equation}
where $Var(.)$ is the variance and $res_{X\sim Y}$ are the residuals in the spline model where $X$ is considered a function of $Y$.
The second new measure introduced in @mbgraphic is aiming to detect also non-functional relations and is based on the idea of distance correlation @szekely2007. (should I repeat definitions here?)
Finally we will also consider the maximal and total information coefficient (MIC and TIC) defined in @Reshef1518 which are based on mutual information found when imposing grids on pair-wise variable plots. These functions are availabel in R via the interface implemented in the minerva package @minerva.
